Simulating an Interplanetary Internet

### 2024-04-10 paper/paper.tex
idk

In the past decade interest in returning back to space has increased
internationally. Many organizations have started sending remote vehicles into
space for the first time~%\cite{(indian space program, spaceX, ect)}
and there are talks of a permanent lunar base being created. With an increased
amount of satellites and potentially permanent bases outside of Earths orbit,
the demand on the interplanetary network (IPN) is increased. The IPN has a
series of unique challenges such as a high amount of error, constantly changing
network topology, and long propagation delay. The solution to these problems is
implementing a delay tolerant network. While space organizations have maintained
the IPN with a set of standard protocols, these may not be sufficient for an
expanding network. The simulator proposed shows how an IPN can handle different
stages of development of an IPN as well as increased throughput requirements.
Before these technological changes can be implemented simulations need to be
created to lower the risk of failure given the high cost of failure. Often
satellites will need to be in operation for years or even decades. The hardware
and software in satellites can often not be updated very well. Thus it will be
very relevant to investigate these issues well before they are needed.

@@

In their 2017 Masters Thesis, Md Monjurul Islam Khan~\cite{Khan2017} from
University of Manitoba wrote on how to create a satellite network simulation
based on NS-3. They used a point to point NS-3 module to create a network of 2
levels of orbiting satellites and fixed points on the earth which acts as
uplinks to the network. The paper analyzed TCP HighSpeed against TCP Vegas slow
start. However this paper stops at expanding the network beyond earth orbit, and
rather focuses on Earth orbit communication.

@@

An interplanetary network is simply a network which where the nodes are
distributed across the surface or in the orbits of different planets or other
celestial objects. In the year of 2024 the current interplanetary is limited to
Mars, Earth, Earth orbit, Moon orbit, and satellites orbiting the sun. The
interplanetary network is also expanded to deep space satellites, like the
Voyager probes. A sample message path across the network could be that a message
is passed from a large network antenna on earth to a earth orbiting satellite,
then when it has direct line of sight with MAVEN (a satellite orbiting mars) it
sends the message. MAVEN then relays the message to a Mars rover like Odyssey.
This description hides the complexity which arises in an interplanetary network
which may seem like a extremely simple network.

In a traditional network a majority of the network traffic will occur over a
wired connection, especially for transfers of large amounts of data. This means
that the error is relatively low. This is no longer the case in the case of
satellites sending messages over potentially millions of kilometers. Atmosphere
in the Earth orbit, solar radiation, and dust cause increased error in sending
the message between entities. Inverse square law dictates that over these large
periods the energy needed to send the message so that it can be read at its
destination. Even the Doppler effect is needed to be accounted for to tune into
radio frequencies on the ISS.~\cite{qslCompensatingDoppler} High error means
that sending a message can no longer be assumed to be sent reliably and has to
be compensated for. Beyond high error, large distances between entities and the
physical limitation that messages cannot travel faster than the speed of light.
Light, or any other energy takes approximately 8 minutes to get from the Sun to
Earth,~\cite{nasa_earth} and thus for a round trip will take 16 minutes to
complete a simple handshake, given that the messages were sent correctly.

Other than physical limitations a interplanetary network has a rapidly changing
topology at anytime. For instance satellites which orbit planets can have their
field of view blocked as they move over the horizon. While terrestrial networks
also have rapidly changing networks at the edges, but the difference is that the
core of the network is often stable. Where depending on the hour, day, year, the
core of the interplanetary network can change drastically. To be able to connect
to the network there are no well known places to access the network, except for
Earth.

@@

To combat these challenges several protocols have been created. The SCPS TP is a
transport protocol which is used on the network layer. UDP is used in the lowest
reliable connections between nodes, and TCP is used in the most reliable network
conditions. When a reliable and efficient method of network transportation is
needed, then SCPS TP is used. It is a modified version of TCP which uses a slow
start protocol, reduced handshakes, reliable data transfer with reduced
confirmations. Other protocols like store and load, and Licklider are used to
also overcome these problems. In this paper since no complete implementation of
SCPS TP has been created for NS-3, New Reno TCP was used instead.

@@

The first assumption made was that all orbits are circular. Realistically,
orbits are oval shaped, with two focal points. A result of this assumption is
the elimination of any difference in orbital speed due to Kepler's second law.
Normally a planet will accelerate or decelerate depending on where it is in its
orbit around the sun. The orbital radius was chosen by taking the average
orbital radius of the real life celestial bodies. For example Earth's average
orbital radius is 149.6 billion meters~\cite{nasa_earth}, and therefore orbits
around the sun in a circular orbit of radius 149.6 billion meters in our model.
Another result of this assumption is that orbits are considered to be centered
around the body which it orbits. The Mars orbit has a focal point that is
further in space, making it more oval shaped than the earth
orbit.~\cite{nasa_mars} While this could make changes to a simulated network in
some very specific situations, the primary behaviour of a solar system is
maintained, like retrograde orbits, periodical alignment, predictable behaviour,
and recursive orbits of sub orbiting object.

@@

Based on these simplifications, a physical simulation with several goals. To
calculate if a line of sight could be made between two communicating entities,
that is if a message can be sent, to calculate the error rate when sending a
message, and to calculate the message propagation delay for the message to be
sent. Line of sight was achieved using simple geometry calculations. All
entities are able to block signals if the ray passes through the radius of the
entity. One of the largest factors contributing to transmission error is the
distance traveled by the electromagnetic signal between transmitting and
receiving nodes. By generalizing Friis' transmission formula\cite{Friis} for
isotropic emitters and receivers, we can obtain the ratio of signal intensity
received to signal intensity transmitted, known as Free Space Path Loss (FSPL).
We can use FSPL as our transmission error rate, giving the following expression:
$Error Rate = {(\frac{4 \pi d}{\lambda})}^2$ where $d$ is the distance between
nodes in metres and $\lambda$ is the wavelength of the electromagnetic wave used
in transmission. As much of space communications utilizes the Ka-band of
frequencies (27--40 GHz)\cite{Morabito_Hastrup} we opted to use a transmission
frequency of 30Ghz. The transmission wavelength can be then calculated according
to $\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light.

@@

The physics simulation modeled the Sun, Earth, Moon, International Space
Station, Mars, and the Mars Orbiter. All were able to connect to the network
except for the the Sun and the Moon. The input of the physics engine is a time
$t$ from some preset initial conditions which outputs the state of all entities.
Which is fed into the NS-3 simulation.

@@

NS-3 is a useful library, but we did run into a number of problems over the
course of our implementation of our simulation. The first problem came when we
first tried to compile the library itself and ran into a bug in the build script
that caused all executables to fail to run if there was the word ``scratch''
anywhere in the path due to a specific subdirectory of NS-3 which also happens
to be called ``scratch''. When first testing it out, one of us had put NS-3 into
a directory that happened to contain that word. We made a pull request to fix
this issue and it was merged by the NS-3 maintainers which allowed us to
continue.

@@

During the work on this project we contributed to the open source NS-3 project 
by committing a change to fix a bug which occurred when running the program on 
different operating systems.

@@

Data was scraped from the trace output that was generated while running the 
ns3 simulation. Trace data was given when a message was queued, dequeued, 
and received at the destination. This data also included the amount of 
data being sent in bytes. There is other data on the messages being sent, but 
was not tracked for the purpose of this project. The trace data was parsed 
using regex expressions to extract the needed numbers from the string.
This data was then fed into a pandas dataframe. A unique id had to be 
generated for each message since once a message was received the id was 
reused. This was done by tracking when each message was received with id 
$x$ and after incrementing a number by one, say $y$. The unique id was then 
created by hashing $xy$. Once the data frame was created it had to be merged 
on itself to have one row of the dataframe represent exactly one message rather 
then one row be either queue, dequeue, or received.

@@

On the left side of the graph before approximately $t = 25000$ this is where
the planet most likely has a direct path between Earth and Mars. This has a
steady pattern of a batch of messages are queued, they are sent, and after they
are sent the next batch of messages are queued, and calculated the propagation
delay at the time that it is queued. This leads to the clusters which can be seen.
The curvature of the delay is most likely attributed to the movement of Earth
relative to Mars, that is that they are getting closer together. Queueing delay is likely small in this section suggested by the tight groupings.

At approximately time $t = 25000$ the direct path between Earth and Mars is
blocked. This means that the messages have to be routed through a secondary
satellite. The vertical gap between the horizontal lines is roughly 0.001
seconds. This is the same time as it takes for a message to take for a message
to send between Earth and the ISS and similarly between Mars and the Mars
Orbiter. The reason for the multiple horizontal lines is that messages may be
passed several times between the planet and its satellite before it is finally
delivered across the large propagation delay between the planets Earth and Mars.
The propagation delay could be seen through the length of the horizontal lines.
This is representative of the time it takes to clear each batch when it is sent.
Once an entire batch is sent it will re-calculate the propagation delay which
will then be represented in new different horizontal lines. The long lines show
that there was a large spike in queuing delay as the topology changed which
became smaller as it cleared. The curvature of this curve likely represents the
change in the relative distance between the satellites orbiting Earth and Mars
becoming further apart. This is clear here due to the extremely small variation
in propagation delay.

This analysis cannot be 100\% verified due to the data mined from NS-3 did not give a ton of information about the messages which were sent on our simulated network.

@@

The average amount of bytes transferred is 320.44 bytes. This is likely from the
difference in the amounts of bytes transferred in an ACK message compared to
when data is being sent.

Unlike in UDP the motion of the planets is not observable in the graph. However
this is because there are several different paths which the TCP message could
have taken which can be seen. Since TCP requires a higher amount of base time to
establish a connection compared to UDP, this means that when it ``hops'' between
entities it adds a larger base time.

@@

New Reno, a modified version of TCP used to help reduce
congestion~\cite{Nahar2016}, was used in place of SCP TP. New Reno behaved very
similar to TCP. Its average data transfer was 307.18 bytes.

It preformed marginally better in its time from being queued to the message
received. This was mostly due to a decrease in time that it was in queue, which
is to be expected. There was a very marginal increase in time in transit
however.

@@

Similar to TCP we can see the two stages which occur in the simulation. One
where most of the messages were likely being routed from directly from Earth to
Mars, which then gets transformed into having to route through one of the
external satellites orbiting each entity, which decreases the average time it
takes to transmit the message.

@@

Since a large amount of propagation delay dominates the amount of time that it
takes for messages total transmit time from being queued to being received at
the appropriate node, other increases in performance mean only a slight decrease
in time sending the message. An important statistic to pay attention to is that
UDP almost always sent a message with a size of 1052 bytes. This means that even
though TCP, New Reno, and UDP had sent a similar number of messages, the actual
throughput for UDP was significantly higher, which is much more significant in
an IPN situation. In the data graphs we can see that not only was the maximum
data being sent was less than UDP, but also had to send messages of very little
size, most likely acknowledgments, as well. New Reno was seen to improve on the
queuing time compared to both TCP and UDP. Which would mean in a high
reliability situation and in situations where higher traffic is sent through a
network with a relatively low bandwidth, like a IPN, where satellites are
limited by how much data they can transmit, New Reno's effect can be more
significant. This should only be used in a connection where propagation delay is
less of an issue like between planet and satellite communication. Queuing time
in UDP took a more significant percentage of its time than both TCP and New
Reno. In general however, UDP is the best option for primary communication
between Mars and Earth communication.

@@

Future work that could be done would be to implement or partially implement SCP
TP as a network protocol which could be used and compared to the protocols
analyzed in this paper. Other protocols like the Linklider or load and store
protocols could also be implemented, to create a more realistic IPN.

@@

In this paper we outlined how we created a model of an interplanetary internet
(IPN) using a physics engine which fed into a NS-3 simulation. The physics
simulation used a simplified model of orbiting planets and satellites as well as
physics calculations to create data which would be the output for any input time
$t$. The NS-3 model used a point to point base which programmatically took the
data from the physics engine and created nodes which acted as routers. A
simulation was ran which transferred data between Earth and Mars, using
satellites and other entities to relay messages around blocked connections.
Analysis was done on three different network protocols used in the IPN, that is
UDP, TCP and New Reno. UDP was found to have the most throughput out of all of
the three protocols, while New Reno reduced the amount of time a message was
queued.

>>>>

In the past decade, interest in returning back to space has increased
internationally. Many organizations have started sending remote vehicles into
space for the first time~%\cite{(indian space program, spaceX, ect)}
and there are talks of a permanent lunar base being created. With an increased
amount of satellites and potentially permanent bases outside of Earths orbit,
the demand on the interplanetary network (IPN) is increased. The IPN has a
series of unique challenges such as a high amount of error, constantly changing
network topology, and long propagation delay. The solution to these problems is
implementing a delay tolerant network. While space organizations have maintained
the IPN with a set of standard protocols, these may not be sufficient for an
expanding network. The simulator proposed shows how an IPN can handle different
stages of development of an IPN as well as increased throughput requirements.
Before these technological changes can be implemented simulations need to be
created to lower the risk of failure given the high cost of failure. Often
satellites will need to be in operation for years or even decades. The hardware
and software in satellites can often not be updated very well. Thus it will be
very relevant to investigate these issues well before they are needed.

@@

In their 2017 Masters Thesis, Md Monjurul Islam Khan~\cite{Khan2017} from
University of Manitoba wrote about how one can create a satellite network
simulation based in NS-3. They used a point to point NS-3 module to create a
network of 2 levels of orbiting satellites and fixed points on the earth which
acts as uplinks to the network. The paper analyzed TCP HighSpeed against TCP
Vegas slow start. However this paper stops at expanding the network beyond earth
orbit, and rather focuses on Earth orbit communication.

@@

An interplanetary network is simply a network which where the nodes are
distributed across the surface or in the orbits of different planets or other
celestial objects. In the year of 2024 the current interplanetary network is
limited to Mars, Earth, Earth orbit, Moon orbit, and satellites orbiting the
sun. The interplanetary network is also expanded to deep space satellites, like
the Voyager probes. A sample message path across the network could be that a
message is passed from a large network antenna on earth to a earth orbiting
satellite, then when it has direct line of sight with MAVEN (a satellite
orbiting mars) it sends the message. MAVEN then relays the message to a Mars
rover like Odyssey. This description hides the complexity which arises in an
interplanetary network which may seem like a extremely simple network.

In a traditional network, a majority of the network traffic will occur over a
wired connection, especially for transfers of large amounts of data. This means
that the error is relatively low. This is no longer the case with satellites
sending messages over potentially millions of kilometers. Atmosphere in the
Earth orbit, solar radiation, and dust cause increased error in sending the
message between entities. Inverse square law dictates that over these large
periods the energy needed to send the message so that it can be read at its
destination. Even the Doppler effect is needed to be accounted for to tune into
radio frequencies on the ISS.~\cite{qslCompensatingDoppler} High error means
that sending a message can no longer be assumed to be sent reliably and has to
be compensated for. Beyond high error, large distances between entities and the
physical limitation that messages cannot travel faster than the speed of light.
Light, or any other energy takes approximately 8 minutes to get from the Sun to
Earth,~\cite{nasa_earth} and thus for a round trip will take 16 minutes to
complete a simple handshake, given that the messages were sent correctly.

Other than physical limitations an interplanetary network has a rapidly changing
topology at all times. For instance, satellites which orbit planets can have
their field of view blocked as they move over the horizon. While terrestrial
networks also have rapidly changing networks at the edges, the core of the
network is often stable. Where depending on the hour, day, year, the core of the
interplanetary network can change drastically. To be able to connect to the
network there are no well known places to access the network, except for Earth.

@@

To combat these challenges several protocols have been created. The SCPS TP is a
transport protocol which is used on the network layer. UDP is used in the lowest
reliable connections between nodes, and TCP is used in the most reliable network
conditions. When a reliable and efficient method of network transportation is
needed, then SCPS TP is used. It is a modified version of TCP which uses a slow
start protocol, reduced handshakes, reliable data transfer with reduced
confirmations. Other protocols like store and load, and Licklider are used to
also overcome these problems. New Reno TCP was used instead of SCPS TP since no
complete implementation of SCPS TP has been created for NS-3.

@@

The first assumption made was that all orbits are circular. Realistically,
orbits are oval shaped, with two focal points. A result of this assumption is
the elimination of any difference in orbital speed due to Kepler's second law.
Normally a planet will accelerate or decelerate depending on where it is in its
orbit around the sun. The orbital radius was chosen by taking the average
orbital radius of the real life celestial bodies. For example, Earth's average
orbital radius is 149.6 billion meters~\cite{nasa_earth} and therefore, it
orbits around the sun in a circular orbit of radius 149.6 billion meters in our
model. Another result of this assumption is that orbits are considered to be
centered around the body which it orbits. The Mars orbit has a focal point that
is further in space, making it more oval shaped than the earth
orbit.~\cite{nasa_mars} While this could make changes to a simulated network in
some very specific situations, the primary behaviour of a solar system is
maintained such as retrograde orbits, periodical alignment, predictable
behaviour, and recursive orbits of sub orbiting object.

@@

We designed a physical simulation with several goals based on these
simplifications. To calculate if a line of sight could be made between two
communicating entities, that is if a message can be sent, to calculate the error
rate when sending a message, and to calculate the message propagation delay for
the message to be sent. Line of sight was achieved using simple geometry
calculations. All entities are able to block signals if the ray passes through
the radius of the entity. One of the largest factors contributing to
transmission error is the distance traveled by the electromagnetic signal
between transmitting and receiving nodes. By generalizing Friis' transmission
formula~\cite{Friis} for isotropic emitters and receivers, we can obtain the
ratio of signal intensity received to signal intensity transmitted, known as
Free Space Path Loss (FSPL). We can use FSPL as our transmission error rate,
giving the following expression: $Error Rate = {(\frac{4 \pi d}{\lambda})}^2$
where $d$ is the distance between nodes in metres and $\lambda$ is the
wavelength of the electromagnetic wave used in transmission. As much of space
communications utilizes the Ka-band of frequencies (27--40
GHz)~\cite{Morabito_Hastrup}, we opted to use a transmission frequency of 30Ghz.
The transmission wavelength can be then calculated according to
$\lambda = \frac{c}{f}$, where $c$ is the speed of light and $f$ is the
transmission frequency, giving a transmission wavelength of about 9.99~mm.
Propagation delay is a simple function of distance and the speed of light.

@@

The physics simulation modeled the Sun, Earth, Moon, International Space
Station, Mars, and the Mars Orbiter. All were able to connect to the network
except for the Sun and the Moon. The input of the physics engine is a time $t$
from some preset initial conditions which outputs the state of all entities.
Which is fed into the NS-3 simulation.

@@

NS-3 is a useful library, but we did run into a number of problems over the
course of our implementation of the simulation. The first problem came when we
first tried to compile the library itself and ran into a bug in the build script
that caused all executables to fail to run if there was the word ``scratch''
anywhere in the path due to a specific subdirectory of NS-3 which also happens
to be called ``scratch''. When first testing it out, one of us had put NS-3 into
a directory that happened to contain that word. We made a pull request to fix
this issue and it was merged by the NS-3 maintainers which allowed us to
continue.

@@

@@

Data was scraped from the trace output that was generated while running the NS-3
simulation. Trace data was given when a message was queued, dequeued, or
received at the destination. This data also included the amount of data being
sent in bytes. There is other data on the messages being sent, but was not
tracked for the purpose of this project. The trace data was parsed using regular
expressions to extract the needed numbers from the string. This data was then
fed into a \texttt{pandas} dataframe. A unique ID had to be generated for each
message since once a message was received the ID was reused. This was done by
tracking when each message was received with ID $x$ and after incrementing a
number by one, say $y$. The unique ID was then created by hashing $xy$. Once the
data frame was created it had to be merged on itself to have one row of the
dataframe represent exactly one message rather then one row be either queue,
dequeue, or received.

@@

On the left side of the graph, before about $t = 25000$ seconds is where there
most likely is a direct path between Earth and Mars. This has a steady pattern
of a batch of messages are queued, they are sent, and after they are sent the
next batch of messages are queued, and calculated the propagation delay at the
time that it is queued. This leads to the clusters which can be seen. The
curvature of the delay is most likely attributed to the movement of Earth
relative to Mars, that is that they are getting closer together. Queuing delay
is likely small in this section suggested by the tight groupings.

At approximately $t = 25000$ seconds, the direct path between Earth and Mars is
blocked. This means that the messages have to be routed through a secondary
satellite. The vertical gap between the horizontal lines is roughly 0.001
seconds. This is the same time as it takes for a message to take for a message
to send between Earth and the ISS and similarly between Mars and the Mars
Orbiter. The reason for the multiple horizontal lines is that messages may be
passed several times between the planet and its satellite before it is finally
delivered across the large propagation delay between the planets Earth and Mars.
The propagation delay could be seen through the length of the horizontal lines.
This is representative of the time it takes to clear each batch when it is sent.
Once an entire batch is sent it will re-calculate the propagation delay which
will then be represented in new different horizontal lines. The long lines show
that there was a large spike in queuing delay as the topology changed which
became smaller as it cleared. The curvature of this curve likely represents the
change in the relative distance between the satellites orbiting Earth and Mars
becoming further apart. This is clear here due to the extremely small variation
in propagation delay.

This analysis cannot be 100\% verified due to the data mined from NS-3 did not
give as much information as we would have liked about the messages which were
sent on our simulated network.

@@

The average amount of bytes transferred is 320.44 bytes. This is likely from the
difference in the amounts of bytes transferred in an ACK message compared to
when data is being sent. On average, the queue delay was negligible. The total
transfer time was 261.56 seconds on average.

Unlike in UDP, the motion of the planets is not observable in the graph. However
this is because there are several different paths which the TCP message could
have taken which can be seen. Since TCP requires a higher amount of base time to
establish a connection compared to UDP, this means that when it ``hops'' between
entities it adds a larger base time.

@@

New Reno, a modified version of TCP used to help reduce
congestion~\cite{Nahar2016}, was used in place of SCP TP. New Reno behaved very
similar to TCP. Its average data transfer was 307.18 bytes. On average, the
queue delay was $8.69\times10^{-4}$ seconds. The average total transfer time was
261.56 seconds. The differences in our simulation between this and TCP were
negligible.

@@

Similar to TCP we can see the two stages which occur in the simulation. One,
where most of the messages were likely being routed directly from Earth to
Mars, which then gets transformed into having to route through one of the
external satellites orbiting each entity, which decreases the average time it
takes to transmit the message.

@@

Since a large amount of propagation delay dominates the total transmit time from
being queued, to being received at the appropriate node, other increases in
performance mean only a slight decrease in time sending the message. An
important statistic to pay attention to is that UDP almost always sent a message
with a size of 1052 bytes. This means that even though TCP, New Reno, and UDP
had sent a similar number of messages, the actual throughput for UDP was
significantly higher, which is much more significant in an IPN situation. In the
data graphs we can see that not only was the maximum data being sent was less
than UDP, but also had to send messages of very little size, most likely
acknowledgments, as well. Queuing time in UDP took a more significant percentage
of its time than both TCP and New Reno. In general however, UDP is the best
option for primary communication between Mars and Earth communication.

@@

Future work that could be done would be to implement or partially implement SCP
TP as a network protocol which could be used and compared to the protocols
analyzed in this paper. The Linklider or store and load protocols could also be
implemented, to create a more realistic IPN.

@@

In this paper, we outlined how we created a model of an interplanetary internet
using a physics engine which fed into a NS-3 simulation. The physics simulation
used a simplified model of orbiting planets and satellites as well as physics
calculations to create data which would be the output for any input time $t$.
The NS-3 model used a point to point base which programmatically took the data
from the physics engine and created nodes which acted as routers. A simulation
was ran which transferred data between Earth and Mars, using satellites and
other entities to relay messages around blocked connections. Analysis was done
on three different network protocols used in the IPN, that is UDP, TCP and New
Reno. UDP was found to have the most throughput out of all of the three
protocols.
